Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2018-11-13 paper/paper.Rmd
made bin functions generalizable and added feature distribution plots

```{r setup, include = FALSE}
# Set knitr options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE)

# Load libraries
library(tidyverse)
library(bulletr)
library(lime)
library(randomForest)
```

@@

@@

```{r}
# Obtain features used when fitting the rtrees random forest
rf_features <- rownames(rtrees$importance)

# If the test data does not exist, clean the data, create the test data, and save it
# Otherwise, load in the test data
if(!file.exists("../data/hamby224_test.csv")) {

@@

Marco Tulio Ribeiro, Sameer Singh, and Carlos Guestrin developed LIME in response to the interpretabiliby issue of many machine learning techniques. Many models used in machine learning are referred to as black box prediction models since they are complex models that may provide accurate predictions but are not able to be interpreted. That is, it is not possible to understand the relationship between the features and the response variable. Without being able to interpret the models, it becomes difficult to diagnose why a model is producing the predictions that it is. The developers claim that LIME allows users to check which variables are driving the predictions made by the model. This further allows users to determine whether their model is trustworthy.

@@

  # Apply the lime function from the lime package (with a seed set)
  # Note that the as_classifier must be added since rtrees is from the
  # randomForest package and not fit using caret or one of the other 
  # available models specified in the lime package. Additionally, the
  # randomForest package must be loaded in order to run the functions
  # from the lime package. (I should check on exactly how this works.)
  set.seed(84902)
  hamby224_lime <- lime(x = hamby173and252_train %>% select(rf_features),
                        model = as_classifier(rtrees))

@@

  # Apply the explain function from the lime package (with a seed set)
  set.seed(84902)
  hamby224_explain <- lime::explain(hamby224_test %>% 
                                      arrange(case) %>% 
                                      select(rf_features) %>% 
                                      na.omit(), 
                                    hamby224_lime,
                                    labels = "FALSE",
                                    n_features = 3)

@@

```{r}
if (!file.exists("../data/hamby_bins.csv")) {

  # Create a dataframe of bins cuts
  hamby_bins <- data.frame(matrix(unlist(hamby224_lime$bin_cuts), 
                                  nrow = 9, byrow = TRUE)) %>%
      rename("Q0" = X1, "Q25" = X2, "Q50" = X3, "Q75" = X4, "Q100" = X5) %>%
      mutate(Feature = rf_features) %>%
      select(Feature, Q0:Q100)

@@

@@

```{r}
# Function to use for creating bin labels in the test_explain dataset
bin_labeller <- function(feature, feature_value, bin_cuts = hamby_bins){

@@

    # Subset the bin cuts table to the selected feature
    feature_bin_cuts <- bin_cuts %>% filter(Feature == feature)

    # Determine which bin the case falls in
    if(feature_value <= feature_bin_cuts$Q25){
      feature_bin <- paste(feature, "(lower bin)")
    } else if (feature_bin_cuts$Q25 < feature_value & 
               feature_value <= feature_bin_cuts$Q50){
      feature_bin <- paste(feature, "(lower middle bin)")
    } else if (feature_bin_cuts$Q50 < feature_value & 
               feature_value <= feature_bin_cuts$Q75){
      feature_bin <- paste(feature, "(upper middle bin)")
    } else if (feature_bin_cuts$Q75 < feature_value){
      feature_bin <- paste(feature, "(upper bin)")
    }

@@

```{r}
# Check to see if model_prediction is the sum of the coefficients from the simple model
hamby224_test_explain %>% 
  filter(case == 1) %>%
  select(model_intercept, feature, feature_weight, model_prediction, rfscore)

simple_model <-  data.frame(term = c("intercept", 
                    as.character(hamby224_test_explain %>% filter(case == 1) %>% pull(feature))),
           coefficients = c(subset(hamby224_test_explain, case == 1)$model_intercept[1],
                            hamby224_test_explain %>% filter(case == 1) %>% pull(feature_weight)))

# It is!
sum(simple_model$coefficients)
subset(hamby224_test_explain, case == 1)$model_prediction[1]
```

@@

THe points are also colored by the $R^2$ value from the simple model. This plot suggests that the simple models with a prediction close to 0.5 have a better fit, but the $R^2$ values are still low. I am not sure what to make of this. However, it does not appear that the $R^2$ is related to the probability from the random forest.

@@

The plots below show the absolute value of the difference between the random forest probability and the simple model probabilty versus the random forest probability. These plots show a clear v-shaped trend that suggests that the simple model is doing the best job of replicating the random forest probabilty around the random forest probability of 0.5. That is, the simple model seems to be doing a worse job of computing accurate predictions in the extremes of the random forest scores.

>>>>

```{r setup, include = FALSE}
# Set knitr options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, fig.pos = "center")

# Load libraries
library(tidyverse)
library(bulletr)
library(randomForest)
library(lime)
library(cowplot)
```

@@

```{r}
# Obtain features used when fitting the rtrees random forest
rf_features <- rownames(rtrees$importance)
```

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Frequency", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
# using the position = "fill" option
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(position = "fill", bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Proportion", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig.height = 4, fig.width = 9}
# Create a correlation heatmap of the match comparisons in the training data
cor_match <- hamby173and252_train %>%
  select(rf_features, samesource) %>%
  filter(samesource == TRUE) %>%
  select(-samesource) %>%
  cor() %>%
  reshape2::melt() %>%
  mutate(Var1 = factor(Var1, levels = c("ccf", "cms", "matches", "rough_cor", "sum_peaks",
                                        "D", "sd_D", "mismatches", "non_cms")),
         Var2 = factor(Var2, levels = c("ccf", "cms", "matches", "rough_cor", "sum_peaks",
                                        "D", "sd_D", "mismatches", "non_cms"))) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() + 
  scale_fill_gradient2(limits = c(-1, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") + 
  labs(x = "", y = "", fill = "Correlation",
       title = "Match Comparisons")

# Create a correlation heatmap of the non-match comparisons in the training data
cor_nonmatch <- hamby173and252_train %>%
  select(rf_features, samesource) %>%
  filter(samesource == FALSE) %>%
  select(-samesource) %>%
  cor() %>%
  reshape2::melt() %>%
  mutate(Var1 = factor(Var1, levels = c("ccf", "cms", "matches", "rough_cor", "sum_peaks",
                                        "D", "sd_D", "mismatches", "non_cms")),
         Var2 = factor(Var2, levels = c("ccf", "cms", "matches", "rough_cor", "sum_peaks",
                                        "D", "sd_D", "mismatches", "non_cms"))) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() + 
  scale_fill_gradient2(limits = c(-1, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "", fill = "Correlation",
       title = "Non-Match Comparisons")

# Create a title for the panel of plots
cor_title <- ggdraw() +
  draw_label("Correlation of Feature Variables in the Training Data", 
             fontface = "bold", 
             size = 16,
             x = 0.5,
             hjust = 0.5)

# Create a joined legend for the panel of plots
cor_legend <- get_legend(cor_match + theme(legend.position = "right"))

# Create the panel of plots
plot_grid(cor_title, 
          plot_grid(cor_match, cor_nonmatch, cor_legend, ncol = 3, rel_widths = c(2, 2, 0.5)),
          ncol = 1,
          rel_heights = c(0.25, 3))
```

@@

```{r}
# If the test data does not exist, clean the data, create the test data, and save it
# Otherwise, load in the test data
if(!file.exists("../data/hamby224_test.csv")) {

@@

Marco Tulio Ribeiro, Sameer Singh, and Carlos Guestrin developed LIME in response to the interpretation issue of many machine learning techniques. Many models used in machine learning are referred to as black box prediction models since they are complex models that may provide accurate predictions but are not able to be interpreted. That is, it is not possible to understand the relationship between the features and the response variable. Without being able to interpret the models, it becomes difficult to diagnose why a model is producing the predictions that it is. The developers claim that LIME allows users to check which variables are driving the predictions made by the model. This further allows users to determine whether their model is trustworthy.

@@

  # Apply the lime function from the lime package (with a seed set)
  # Note that the as_classifier must be added since rtrees is from the
  # randomForest package and not fit using caret or one of the other 
  # available models specified in the lime package. Additionally, the
  # randomForest package must be loaded in order to run the functions
  # from the lime package. (I should check on exactly how this works.)
  set.seed(84902)
  hamby224_lime <- lime(x = hamby173and252_train %>% select(rf_features),
                        model = as_classifier(rtrees),
                        n_bins = 6)

@@

  # Apply the explain function from the lime package (with a seed set)
  set.seed(84902)
  hamby224_explain <- lime::explain(hamby224_test %>% 
                                      arrange(case) %>% 
                                      select(rf_features) %>% 
                                      na.omit(), 
                                    hamby224_lime,
                                    labels = "TRUE",
                                    n_features = 3)

@@

```{r}
# Function for writing bins given the bin boundaries
write_bins <- function(bin_data){

  # Compute the number of bins based the input dataframe
  nbins = length(bin_data) - 2

  # Create an empty matrix to store the bins in
  row <- matrix(NA, nrow = 1, ncol = nbins + 1)

  # Put the feature name in the first column
  row[,1] <- as.character(bin_data[[1]])

  # Put the 1st through 2nd to last bins in the 2nd through 2nd to last columns
  row[,2:nbins] <- sapply(2:nbins, 
                          FUN = function(number) sprintf("[%.2f, %.2f)", 
                                                         as.numeric(bin_data[number]),
                                                         as.numeric(bin_data[number + 1])))

  # Fill in the last column with the upper bin (based on whether it should be infinite or not)
  row[,nbins + 1] <- ifelse(is.na(bin_data[nbins + 2]),
                            sprintf("(%.2f, %s)", as.numeric(bin_data[nbins + 1]), "\U221E"),
                            sprintf("(%.2f, %.2f]", as.numeric(bin_data[nbins + 1]),
                                    as.numeric(bin_data[nbins + 2])))

  return(row)

```{r}
# Create a dataframe with the bins
if (!file.exists("../data/hamby_bins.csv")) {

  # Determine the nubmer of bins
  nbins <- length(hamby224_lime$bin_cuts[[1]]) - 1

  # Create a dataframe of bin boundaries
  hamby_divisions <- data.frame(Feature = rf_features, 
                                Lower = c(0, -1, 0, 0, 0, 0, 0, 0, 0),
                           matrix(unlist(hamby224_lime$bin_cuts), nrow = 9, 
                                  byrow = TRUE)[,2:nbins],
                           Upper = c(1, 1, rep(NA, 7)))

  # Use my function "write_bins" to create a dataframe with the bins
  hamby_bins <- data.frame(t(apply(hamby_divisions, 1, write_bins)))

  # Assign appropriate names to the bin columns
  names(hamby_bins) <- c("Feature", "Lower Bin", 
                         sapply(1:(nbins-2), function(bin) sprintf("Middle Bin %.0f", bin)),
                         "Upper Bin")

@@

  # Replace the periods with spaces in the feature names
  names(hamby_bins) <- gsub(".", " ", names(hamby_bins), fixed = TRUE)

@@

```{r}
# Function to use for creating bin labels in the test_explain dataset
bin_labeller <- function(feature, feature_value, bin_data = hamby_divisions){

@@

    # Subset the bin cuts table to the selected feature
    feature_bin_data <- bin_data %>%
      filter(Feature == feature) %>%
      select(-Feature, -Lower, -Upper)

    # Compute the number of bins based the input dataframe
    nbins = length(bin_data) - 2

    # Determine which bin the case falls in
    if(feature_value <= feature_bin_data[1]){
      feature_bin <- paste(feature, "(lower bin)")
    } else if (feature_bin_data[nbins - 1] < feature_value){
      feature_bin <- paste(feature, "(upper bin)")
    } else {
      middle_bin_checks <- data.frame(middle_bin_number = 1:(nbins-2),
                                      contained = sapply(1:(nbins - 2),
                                                         FUN = function(number) {
                                                           feature_bin_data[number] < feature_value &
                                                           feature_value <= feature_bin_data[number + 1]}))
      feature_bin <- sprintf("%s (middle bin %.0f)", 
                             feature,
                             middle_bin_checks %>% filter(contained == TRUE) %>% pull(middle_bin_number))
    }

@@

@@

The points are also colored by the $R^2$ value from the simple model. This plot suggests that the simple models with a prediction close to 0.5 have a better fit, but the $R^2$ values are still low. I am not sure what to make of this. However, it does not appear that the $R^2$ is related to the probability from the random forest.

@@

The plots below show the absolute value of the difference between the random forest probability and the simple model probability versus the random forest probability. These plots show a clear v-shaped trend that suggests that the simple model is doing the best job of replicating the random forest probability around the random forest probability of 0.5. That is, the simple model seems to be doing a worse job of computing accurate predictions in the extremes of the random forest scores.
